Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-07 manuscript/climate_sensitivity_draft.Rmd
incorporated (most of) JZ's edits

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
2.
3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
4.
5.
6.
7.
8.

@@

### Summary
- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important. Yet it remains unclear how tree size and species' traits interactively shape tree growth responses across droughts.

@@

- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance. Resistance was lower among species whose leaves shrunk more upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across given droughts.

@@

\newpage
### Introduction
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies in responses to drought (*[Kennedy et al. 2019](https://doi.org/10.1029/2018MS001500)*). In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], even in conjunction with increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Global change-type drought has been affecting forests worldwide [@allen_global_2010], and it is expected that future climate change-driven droughts will severely impact forests around the world [@allen_global_2010]; **REFS**). Larger trees tend to suffer more (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (*e.g., McDowell et al. 2016(doi:10.1038/nclimate2873); Charney et al. 2016 (doi: 10.1111/ele.12650)*), yet accurate characterization of drought responses remains a modeling challenge (*[Kennedy et al. 2019](https://doi.org/10.1029/2018MS001500)*)-- in part because some of the mechanisms underlying drought responses remain unclear. Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], and this may become a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; **REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response. Xylem architecture plays a role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but does not differentiate species beyond broad classes. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018). Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019]. The ability of these hydraulic traits to explain tree performance under drought remains untested.

A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts. No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018]. However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (*e.g.*, **REFS??**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.

@@

Hydraulic traits were collected at SCBI (Table 3). In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot. These were covered with opaque plastic bags, re-cut under water, and re-hydrated overnight before further analysis. Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using an R script (**details - Valentine? Nobby?**). $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA$ was calculated as the percent loss of area between fresh and dry leaves. $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume. We used the rapid determination method of @bartlett_rapid_2012 to estimate the turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012]. We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration. Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].

To characterize how environmental conditions vary with height, data were obtained from the a National Ecological Observation Network (NEON) tower located <1km from the study area. We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018]. After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included.

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).

@@

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**). There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was correlated with the hydraulic leaf traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (@greenwood_tree_2017; @anderegg_meta-analysis_2016). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response. Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. Vertical gradients in stem and leaf hydraulic traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges. Taller trees also face different microenvironments (Fig. 2a-b), in part because they are more likely to be in dominant canopy positions (Fig. 2d). Even under non-drought conditions, evaporative demand increases with tree height, and tall trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height. Correlation between height and canopy position (Fig. 2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment. However, signficant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

>>>>

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
2.
3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
4.
5. Biological Sciences Department; California State University; Los Angeles, CA, USA
6.
7.
8.

@@

### Summary
- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important. Yet it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.

@@

- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance. Resistance was lower among species whose leaves had higher percent loss of area upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across given droughts.

@@

\newpage
### Introduction
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies in responses to drought (*[Kennedy et al. 2019](https://doi.org/10.1029/2018MS001500)*). In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Global change-type droughts have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests around the world [@allen_global_2010]; **REFS**). During drought, larger trees generally suffer greater reductions in growth and increases in mortality (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (*e.g., McDowell et al. 2016(doi:10.1038/nclimate2873); Charney et al. 2016 (doi: 10.1111/ele.12650)*). Yet, accurate characterization of drought responses remains a modeling challenge (*[Kennedy et al. 2019](https://doi.org/10.1029/2018MS001500)*)-- in part because some of the mechanisms underlying drought responses remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], which becomes a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; **REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response. Xylem architecture plays a role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but does not differentiate species beyond broad classes. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018). Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the osmotic potential at turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. The ability of $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether variability in factors such as drought severity, duration, and timing interact with tree size and traits such that components of the community respond differently across droughts. No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018]. However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (*e.g.*, **REFS??**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019]. [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.

@@

Hydraulic traits were collected at SCBI (Table 3). In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot. Samples were re-cut under water by at least two notes and re-hydrated overnight in buckets covered with opaque plastic bags before measurements. Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the Leafarea R package (**citation**). $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA$ was calculated as the percent loss of area between fresh and dry leaves. $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume. We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012]. Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].

To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area. We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018]. After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\le$ 2) in the univariate models but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included. $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ and $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).

@@

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**). There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response. Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges. Taller trees also face different microenvironments (Fig. 2a-b), in part because they are more likely to be in dominant canopy positions (Fig. 2d). Even under non-drought conditions, evaporative demand increases with tree height in taller trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (*[Smith & Nobel 1977](https://doi.org/10.1093/jxb/28.1.169)*), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height. Correlation between height and canopy position (Fig. 2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment. However, signficant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
